Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we will to introduce a \textit{buffer} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. On the other hand, reads events will first attempt
to read from the buffer, and, if there is no corresponding write in the buffer,
will read an arbitrary value, in both cases pushing an event to the global
pomset. At any point, we may choose to \textit{flush} some of the writes in
buffer to the global pomset of events. In particular, we will introduce the set
\(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write actions}, where an
action \(\bufloc{x} := v\) by a thread denotes adding a write \(x := v\) to the
thread's write buffer. We may then define the set of TSO actions
\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be defined
to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret
as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index. In particular, we
define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
where \(\Sigma\) is the stream action of finite pomsets over
\(\mc{A}_{\ms{TSO}}\) on pomsets over \(\mc{A}_{\ms{TSO}}\); we can view
\(\ms{PO}\) as a submonad of this monad in the standard manner.

Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! The semantics of reads and writes are then given
by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
\cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}

\TODO{fence}

\TODO{big power, such transactional memory, much wow}

We can now define the parallel composition of morphisms in a "fork-join" style as follows: we first flush the buffer completely (i.e., taking it and sticking it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel with separate buffers, \textit{filtering out executions which completely flush the buffer}. Since both threads end up with an empty buffer, the resulting joined buffer is also empty, giving us the following:
\begin{equation}
  \begin{aligned}
    f_0 || f_1 = \lambda (a_0, a_1)\;L. 
    & \{\iota_0 ((b_0, b_1), [], L;(\alpha_0 || \alpha_1)) 
      \mid \iota_0 (b_i, [], \alpha_i) \in f_i\;a_i\;[]\} 
    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, [], \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0\;[]) \land \iota_1 \alpha_1 \in f_1\;a_1\;[]\} 
    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0\;[] \land \iota_0 (a_1, [], \alpha_1) \in f_1\;a_1\;[]\}
    \end{aligned}
\end{equation}

>>>>

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we will to introduce a \textit{buffer} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. On the other hand, reads events will first attempt
to read from the buffer, and, if there is no corresponding write in the buffer,
will read an arbitrary value, in both cases pushing an event to the global
pomset. At any point, we may choose to \textit{flush} some of the writes in
buffer to the global pomset of events. In particular, we will introduce the set
\(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write actions}, where an
action \(\bufloc{x} := v\) by a thread denotes adding a write \(x := v\) to the
thread's write buffer. We may then define the set of TSO actions
\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be defined
to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret
as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index (with the empty list corresponding to \(\{\delta\}\)). In particular, we
define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
where \(\Sigma\) is the stream action of finite pomsets over
\(\mc{A}_{\ms{TSO}}\) on pomsets over \(\mc{A}_{\ms{TSO}}\); we can view
\(\ms{PO}\) as a submonad of this monad in the standard manner.

Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! The semantics of reads and writes are then given
by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
\cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
To be able to perform synchronization, we will also need to introduce a \(\ms{fence}: \mc{I}^\varnothing_0(\mb{1}, \mb{1})\) instruction, which simply causes all actions before the fence to be observed before any actions after the fence. For \(\ms{TSO}\), implementing this is as simple as flushing the buffer, giving
\begin{equation}
  \dnt{\ms{fence}} = \lambda ()\;L. \{((), [], L;\{\delta\})\}
\end{equation}
We can now define the parallel composition of morphisms in a "fork-join" style as follows: we first flush the buffer completely (i.e., taking it and sticking it at the beginning of our pomset), then execute \(f\) and \(g\) in parallel with separate buffers, \textit{filtering out executions which completely flush the buffer}. Since both threads end up with an empty buffer, the resulting joined buffer is also empty, giving us the following:
\begin{equation}
  \begin{aligned}
    f_0 || f_1 = \lambda (a_0, a_1)\;L. 
    & \{\iota_0 ((b_0, b_1), [], L;(\alpha_0 || \alpha_1)) 
      \mid \iota_0 (b_i, [], \alpha_i) \in f_i\;a_i\;[]\} 
    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid (\iota_0 (b_0, [], \alpha_0) \in f_0\;a \lor \iota_0\;\alpha_0 \in f_0\;a_0\;[]) \land \iota_1 \alpha_1 \in f_1\;a_1\;[]\} 
    \\ & \cup \{\iota_1 (\alpha_0 || \alpha_1) \mid \iota_1\;\alpha_0 \in f_0\;a_0\;[] \land \iota_0 (a_1, [], \alpha_1) \in f_1\;a_1\;[]\}
    \end{aligned}
\end{equation}
